The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-26 article_final/main.tex
added preliminaries

% Inner product
\newcommand{\<}{\langle}
\renewcommand{\>}{\rangle}

@@

Training a neural network is essentially an optimization of a function of several weight matrices and a few vectors. Let us start by considering the problem of minimizing a differentiable function of a {\it single} matrix:
\begin{equation}\label{eq:matrix_problem}
	F(\cdot)\colon \Rmn \to \R\,,\qquad F(\mX) \to \min_{\mX \in \Rmn}\,
\end{equation}
We equip the matrix space $\Rmn$ with a standard dot product $\<\cdot, \cdot> \to \R$ and a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which does not have to coincide with the Frobenius norm $\normf{\cdot}$. The dual norm $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ that is associated with $\norm{\cdot}$ is defined as
\begin{equation}\label{eq:dual_norm}
	\norm{\mG}^\dagger = \sup_{\mD \in \Rmn\,:\norm{\mD}\leq 1} \<\mG,\mD>\,.
\end{equation}

@@

$$
\cS = \cB_{\norm{\cdot}} = \{ \mD \in \Rmn \mid \norm{\mD} \leq 1 \}\,.
$$
In this case,
$$
\argmin_{\mD \in \cS} \<\mM^k, \mD> = - \{ \mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}\,,
$$
and update for $\mX^k$ in \cref{eq:simple_update} simplifies to
\begin{equation}\label{eq:our_update}
	\mX^{k+1} = \mX^{k} - \gamma_k \{\mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}\,.
\end{equation}

@@

For example, when the norm $\norm{\cdot}$ is the Frobenius norm $\normf{\cdot}$, \cref{eq:our_update} turns into
\begin{equation}\label{eq:nsgd_update}
	\mX^{k+1} = \mX^{k} - \eta_k \frac{\mM^k}{\normf{\mM^k}}\,,
\end{equation}
which recovers Normalized SGD (NSGD).

And when the norm is the spectral norm $\norms{\cdot}$, we get
\begin{equation}\label{eq:muon_update}
	\mX^{k+1} = \mX^{k} - \eta_k U V^\top\,,
\end{equation}
which is Muon without the $\sqrt{m/n}$ factor.
Here, $\mM^k = \mU \mSigma \mV^\top$ is the Singular Value Decomposition (SVD) of $\mM^k$ ($\mU = [u_1, u_2, \dots, u_r]$, $\mSigma = \diag(\sigma_1, \space \sigma_2, \dots, \sigma_r)$, and $\mV = [v_1, v_2, \dots, v_r]$). Muon can be recovered by taking the RMS-to-RMS operator norm $\sqrt{\frac{n}{m}}\norms{\cdot}$.

>>>>

% Inner product notation
\def\<#1,#2>{\langle #1,#2\rangle}

@@

Training a neural network is essentially an optimization of a function of several weight matrices and a few vectors. Let us start by considering the problem of minimizing a differentiable function of a \emph{single} matrix:
\begin{equation}\label{eq:matrix_problem}
	F(\cdot)\colon \Rmn \to \R\,,\qquad F(\mX) \to \min_{\mX \in \Rmn}\,
\end{equation}
We equip the matrix space $\Rmn$ with a standard dot product $\<\cdot, \cdot> \to \R$ and a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which does not have to coincide with the Frobenius norm $\normf{\cdot}$. The dual norm $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ that is associated with $\norm{\cdot}$ is defined as
\begin{equation}\label{eq:dual_norm}
	\norm{\mG}^\dagger = \sup_{\mD \in \Rmn\,:\norm{\mD}\leq 1} \<\mG,\mD>\,.
\end{equation}

@@

\[
\cS = \cB_{\norm{\cdot}} = \{ \mD \in \Rmn \mid \norm{\mD} \leq 1 \}\,.
\]
In this case,
\[
\argmin_{\mD \in \cS} \<\mM^k, \mD> = - \{ \mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}\,,
\]
and update for $\mX^k$ in \equationref{eq:simple_update} simplifies to
\begin{equation}\label{eq:our_update}
	\mX^{k+1} = \mX^{k} - \gamma_k \{\mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}\,.
\end{equation}

@@

For example, when the norm $\norm{\cdot}$ is the Frobenius norm $\normf{\cdot}$, \equationref{eq:our_update} turns into
\begin{equation}\label{eq:nsgd_update}
	\mX^{k+1} = \mX^{k} - \eta_k \frac{\mM^k}{\normf{\mM^k}}\,,
\end{equation}
which recovers Normalized SGD (NSGD).

And when the norm is the spectral norm $\norms{\cdot}$, we get
\begin{equation}\label{eq:muon_update}
	\mX^{k+1} = \mX^{k} - \eta_k \mU \mV^\top\,,
\end{equation}
which is Muon without the $\sqrt{m/n}$ factor.
Here, $\mM^k = \mU \mSigma \mV^\top$ is the Singular Value Decomposition (SVD) of $\mM^k$ ($\mU = [u_1, u_2, \dots, u_r]$, $\mSigma = \diag(\sigma_1, \sigma_2, \dots, \sigma_r)$, and $\mV = [v_1, v_2, \dots, v_r]$). Muon can be recovered by taking the RMS-to-RMS operator norm $\sqrt{\frac{n}{m}}\norms{\cdot}$.
